---
title: "Automate Uninstall, detach, install and attaching of packages"

output: html_document
---

<style type="text/css">

body, td {
   font-family: Segoe UI Light;
   font-size: 20px;
}
</style>

#### In this Blog I will be sharing simple commands to automate Uninstall, Attach, Detach and Install of packages

Here are some smart ways to make your code concise, less error-prone and automate some of the activities in a beautified coding manner. Let me take an example of detaching, uninstalling and installing a package `rvest` .

### Detaching the package


```{r  warning=FALSE}

# Let us first see what are the packages attached in current session

search()

# As we can see that rvest is currently attached. Let us detach it using the following command

detpack = c("package:rvest")

if (detpack %in% search()) { detach(detpack, unload=TRUE,character.only = TRUE) }

```

`rvest` package has been detached successfully.

## Uninstalling the package

In order to update a package, it has to be uninstalled first before installing the fresh version of package. Here we will uninstall `rvest` package using some simple effective commands

```{r  message=FALSE, warning=FALSE}

# First step is to get the list of installed packages which can be retrieved using installed.packages() command
#
# Let us see the sample output of the command
#
head(installed.packages())
#
# Check if rvest is among the installed packages, remove it
#

if ("rvest" %in% rownames(installed.packages())) { remove.packages("rvest") }

```


`rvest` package has been successfully uninstalled.


## Installing and initialize the package

Here we will uninstall and initialize `rvest` package using following commands

```{r}

# Now we download and install rvest for R.

EnsurePackage<-function(x)
{
  x<-as.character(x)
  #
  if (!require(x,character.only=TRUE))   # character.only tells require that x should have only character string
  {
    install.packages(pkgs=x,dependencies = TRUE)
    require(x,character.only=TRUE)
  }
}

EnsurePackage("rvest")

#


```


## I hope you like this clean way of detaching, uninstalling, installing and initialize packages in R to make your program standardized and modularized.
